
Load Libraries 
```{r}
library(tidyverse)
library(dplyr)
library(sf)
library(openxlsx)
library(janitor)
library(htmlwidgets)
library(plotly)
library(tmap)
library(terra)
library(exactextractr)
library(arrow)
library(sfarrow)
```

# Load Data 

Load Energy Data & Filter for TEXAS 
```{r}
utility_TX <- read.xlsx(here::here("data", "raw", "eia8602024", "1___Utility_Y2024.xlsx"), startRow = 2) %>% clean_names() %>% filter(state == "TX")

plant_TX <- read.xlsx(here::here("data", "raw", "eia8602024", "2___Plant_Y2024.xlsx"), startRow = 2) %>% clean_names() %>% filter(state == "TX") %>% rename(cnty_name = county)

wind_energy_TX <- read.xlsx(here::here("data", "raw", "eia8602024", "3_2_Wind_Y2024.xlsx"), startRow = 2) %>% clean_names() %>% filter(state == "TX") %>% rename(cnty_name = county)

solar_energy_TX <- read.xlsx(here::here("data", "raw", "eia8602024", "3_3_Solar_Y2024.xlsx"), startRow = 2) %>% clean_names() %>% filter(state == "TX") %>% rename(cnty_name = county)

storge_TX <- read.xlsx(here::here("data", "raw", "eia8602024", "3_4_Energy_Storage_Y2024.xlsx"), startRow = 2) %>% clean_names() %>% filter(state == "TX") %>% rename(cnty_name = county)

generator_TX <- read.xlsx(here::here("data", "raw", "eia8602024", "3_1_Generator_Y2024.xlsx"), startRow = 2) %>% clean_names() %>% filter(state == "TX") %>% rename(cnty_name = county)

multifuel_TX <- read.xlsx(here::here("data", "raw", "eia8602024", "3_5_Multifuel_Y2024.xlsx"), startRow = 2) %>% clean_names() %>% filter(state == "TX") %>% rename(cnty_name = county)
```

Load Texas Socio-demographic 
```{r}
#st_layers(here::here("data", "raw", "ACS_2019_5YR_TRACT_48_TEXAS.gdb"))

# Geom 
geom_texas <- st_read(here::here("data", "raw", "ACS_2019_5YR_TRACT_48_TEXAS.gdb"), layer = "ACS_2019_5YR_TRACT_48_TEXAS")%>% clean_names()

# Socio-demo info 
age_sex_texas <- st_read(here::here("data", "raw", "ACS_2019_5YR_TRACT_48_TEXAS.gdb"), layer = "X01_AGE_AND_SEX")%>% clean_names()
income_texas <- st_read(here::here("data", "raw", "ACS_2019_5YR_TRACT_48_TEXAS.gdb"), layer = "X19_INCOME")%>% clean_names()
race_texas <- st_read(here::here("data", "raw", "ACS_2019_5YR_TRACT_48_TEXAS.gdb"), layer = "X02_RACE")%>% clean_names()
poverty_texas <- st_read(here::here("data", "raw", "ACS_2019_5YR_TRACT_48_TEXAS.gdb"), layer = "X17_POVERTY")%>% clean_names()
education_texas <- st_read(here::here("data", "raw", "ACS_2019_5YR_TRACT_48_TEXAS.gdb"), layer = "X15_EDUCATIONAL_ATTAINMENT")%>% clean_names()

# Meta data 
metadata_socio_texas <- st_read(here::here("data", "raw", "ACS_2019_5YR_TRACT_48_TEXAS.gdb"), layer = "TRACT_METADATA_2019")
```

Load Texas counties 
```{r}
#texas_counties_boundaries <- st_read(here::here("data", "raw", "Texas_County_Boundaries"))%>% clean_names() # Not what I want 

texas_counties <- st_read(here::here("data", "raw", "tl_2024_us_county"))%>% clean_names() %>% 
  filter(statefp == '48') %>% # THIS IS WHAT I WANT 
  rename(cnty_name = name) # Rename county name colunm to `cnty_name`
```

Load Texas environmental 
```{r}
# GHI data 
sunshine <- terra::rast(here::here("data", "raw", "nsrdbv3_ghi", "Annual GHI", "nsrdb3_ghi.tif"))%>% clean_names()

# Wind data
wind_envir80 <- terra::rast(here::here("data", "raw", "us-wind-data", "wtk_conus_80m_mean_masked.tif"))%>% clean_names()
wind_envir100 <- terra::rast(here::here("data", "raw", "us-wind-data", "wtk_conus_100m_mean_masked.tif"))%>% clean_names()
wind_envir120 <- terra::rast(here::here("data", "raw", "us-wind-data", "wtk_conus_120m_mean_masked.tif"))%>% clean_names()
wind_envir140 <- terra::rast(here::here("data", "raw", "us-wind-data", "wtk_conus_140m_mean_masked.tif"))%>% clean_names()

# Coal deposites 
coal_dep <- read_csv(here::here("data", "raw", "coal_deposites", "TX_points.csv")) %>% clean_names() %>% filter(state_cd == "TX")
```

**The wind data**
This data has wind speed measured at a different height above ground (in meters), ranging from 10 to 200m.
Taller = faster winds because surface friction slows wind near the ground.

10m — Surface weather, not useful for energy. 
80–100m — Onshore wind turbine hub height, most common for wind energy analysis. 
120–140m — Modern large turbines (increasingly common in West Texas). 
200m — Offshore or research purposes. 

Most of the existing West Texas wind farms were built around 80m hub height. New wind turbine development is reaching heights of 100–120m. For this database, I will create a data frame with all possible wind turbine heights ranging from 80m to 140m. 

# Wrangling 

## Socio-demographic Data

Making socio-demographic dataframe spatial 
Join socio-demographic dataframe  with `geom_texas` to add the countyfp column and make spatial 
```{r}
income_texas_sf <- dplyr::left_join(geom_texas, income_texas, by = c("geoid_data"= "geoid")) 
race_texas_sf <- dplyr::left_join(geom_texas, race_texas, by = c("geoid_data"= "geoid")) 
poverty_texas_sf <- dplyr::left_join(geom_texas, poverty_texas, by = c("geoid_data"= "geoid"))
education_texas_sf <- dplyr::left_join(geom_texas, education_texas, by = c("geoid_data"= "geoid"))
age_sex_texas_sf <- dplyr::left_join(geom_texas, age_sex_texas, by = c("geoid_data"= "geoid"))

# Remove non spatial dfs 
#rm(list = c('income_texas', 'race_texas', 'poverty_texas', 'education_texas', 'age_sex_texas'))
```

Per county 
Pull out important columns 
```{r}
income_texas_county <- income_texas_sf %>% 
  group_by(countyfp) %>% 
  summarise(income_avg = mean(b19001e1, na.rm = TRUE))

race_texas_county <- race_texas_sf %>% 
  group_by(countyfp) %>% 
  summarise(pop_total = sum(b02001e1, na.rm = TRUE), 
            white_total = sum(b02001e2, na.rm = TRUE), 
            black_aa_total = sum(b02001e3, na.rm = TRUE), # Blacks / African American 
            ai_an_total = sum(b02001e4, na.rm = TRUE), # American India + Alaska Native
            asian_total = sum(b02001e5, na.rm = TRUE), 
            ah_pi_total = sum(b02001e6, na.rm = TRUE), # Native Hawian, other pacific islander
            other_total = sum(b02001e7, na.rm = TRUE),
            `2plus_total` = sum(b02001e8, na.rm = TRUE)) # two or more races 

poverty_texas_county <- poverty_texas_sf %>% 
  group_by(countyfp) %>% 
  summarise(hh_pov_status_total = sum(b17017e1, na.rm = TRUE))# households Counts in poverty 


education_texas_county <- education_texas_sf %>% 
  group_by(countyfp) %>% 
  summarise(edu_total = sum(b15001e1, na.rm = TRUE))#  EDUCATIONAL ATTAINMENT FOR THE POPULATION 18 YEARS AND OVER 

age_sex_texas_county <- age_sex_texas_sf %>% 
  group_by(countyfp) %>% 
  summarise(male_total = sum(b01001e2, na.rm = TRUE), 
            female_total = sum(b01001e26, na.rm = TRUE), 
            median_age_mean = mean(b01002e1, na.rm = TRUE), 
            median_male_age_mean = mean(b01002e2, na.rm = TRUE), 
            median_female_age_mean = mean(b01002e3, na.rm = TRUE))
  
```


Needs to be per-capita data 
Merge all datasets together 
```{r}
# Make the COUNTYFP a numeric and order the lists based on it 
race_texas_county <- race_texas_county %>% arrange(as.numeric(countyfp))
income_texas_county <- income_texas_county %>% arrange(as.numeric(countyfp))
poverty_texas_county <- poverty_texas_county %>% arrange(as.numeric(countyfp))
education_texas_county <- education_texas_county %>% arrange(as.numeric(countyfp))
age_sex_texas_county <- age_sex_texas_county %>% arrange(as.numeric(countyfp))

# Merge the sociodemographic data 
final_df_sf <- cbind(race_texas_county, income_texas_county, poverty_texas_county, education_texas_county, age_sex_texas_county) 

# Remove unnecessary columns 
final_df_sf <- final_df_sf %>% 
  select(-c(`countyfp.1`, `countyfp.2`, `countyfp.3`, `countyfp.4`, `Shape.1`, `Shape.2`, `Shape.3`, `Shape.4`))

```

Pre capita 
```{r}
# Divide the columns by the population column 
final_df_sf[, 3:17] <- final_df_sf[, 3:17] / final_df_sf$pop_total
```

Add the county names to the data 
```{r}
# Drop the geometry from the socio demographic data 
final_df <- final_df_sf %>% 
  st_drop_geometry()

# Join sociodemo with TX counties  
socio_demo_texas <- left_join(final_df, texas_counties, by = "countyfp")

# Make spatial 
socio_demo_texas <- st_as_sf(socio_demo_texas)
```

Looking at outputs 
`socio_demo_texas` should be 254 rows (254 counties in TX)
```{r}
dim(socio_demo_texas)
dim(texas_counties)
dim(final_df)
n_distinct(socio_demo_texas$countyfp)
```

Create map to show results and dataframes 
```{r}
# TX_counties_static <- ggplot(texas_counties) + 
#   geom_sf(aes(text = paste0("County name:", {name}, "<br>", 
#                             "countyfp:", {countyfp}, "<br>",
#                             "geoid:", {geoid}, "<br>"
#                             )))
# 
# final_df_static <- ggplot(final_df_sf) + 
#   geom_sf(aes(text = paste0("countyfp:", {countyfp}, "<br>"
#                             )))
# 
# 
# socio_demo_static <- ggplot(socio_demo_texas) + 
#   geom_sf(aes(text = paste0("County name:", {name}, "<br>", 
#                             "countyfp:", {countyfp}, "<br>" )))
# 
# TX_counties_static
# final_df_static
# socio_demo_static
# 
# ggplotly(TX_counties_static, tooltip = "text")
# ggplotly(final_df_static, tooltip = "text")
# ggplotly(socio_demo_static, tooltip = "text")

# socio_demo <- ggplot(socio_demo_texas) + 
#   geom_sf(aes(fill = ai_an_total,
#               text = paste0("County name:", {cnty_name}, "<br>", 
#                             "countyfp:", {countyfp}, "<br>",
#                             "geoid:", {geoid}, "<br>"
#                             )))
# 
# ggplotly(socio_demo, tooltip = "text")
```

## Energy data 

Clean / remove unnessary columns  
```{r}
utility_TX <- utility_TX %>% 
  select(-c(owner_of_plants_reported_on_form, operator_of_plants_reported_on_form, asset_manager_of_plants_reported_on_form, other_relationships_with_plants_reported_on_form))

plant_TX <- plant_TX %>% 
  select(-c(latitude, longitude, name_of_water_source, primary_purpose_naics_code,ferc_cogeneration_status, ferc_cogeneration_docket_number, ferc_small_power_producer_status, ferc_small_power_producer_docket_number, ferc_small_power_producer_status, ferc_exempt_wholesale_generator_status, ferc_exempt_wholesale_generator_docket_number, ash_impoundment, ash_impoundment_lined, ash_impoundment_status, transmission_or_distribution_system_owner, energy_storage, natural_gas_ldc_name, natural_gas_pipeline_name_1, natural_gas_pipeline_name_2, natural_gas_pipeline_name_3,pipeline_notes, natural_gas_storage, liquefied_natural_gas_storage, transmission_or_distribution_system_owner_state))

generator_TX <- generator_TX %>% 
  select(c(utility_id, utility_name, plant_code, plant_name, state, cnty_name, generator_id, technology, prime_mover, ownership, nameplate_capacity_mw, nameplate_power_factor, summer_capacity_mw, winter_capacity_mw, minimum_load_mw, status, operating_month, operating_year, sector_name, sector, energy_source_1, multiple_fuels))
       
wind_energy_TX <- wind_energy_TX %>% 
  select(-c(predominant_turbine_model_number, predominant_turbine_manufacturer))
 
solar_energy_TX <- solar_energy_TX %>% 
  select(-c(lenses_mirrors, single_axis_tracking, dual_axis_tracking, fixed_tilt, bifacial,east_west_fixed_tilt, parabolic_trough, linear_fresnel, power_tower, dish_engine, other_solar_technology, azimuth_angle, tilt_angle, crystalline_silicon, thin_film_cd_te, thin_film_a_si, thin_film_cigs, thin_film_other, other_materials))

storge_TX <- storge_TX[, 1:20]

multifuel_TX <- multifuel_TX %>% 
  select(-c(cofire_energy_source_1, cofire_energy_source_2, cofire_energy_source_3, cofire_energy_source_4, cofire_energy_source_5, cofire_energy_source_6, time_to_switch_from_gas_to_oil, factors_that_limit_switching, storage_limits, other_limits))

```



## Sunshine Data 

Crop/Mask to TX 
```{r}
# Look at data 
# Its South and North America
tm_shape(sunshine) + 
  tm_raster()

# Make CRS match 
texas_counties_proj <- st_transform(texas_counties, crs(sunshine))

# Crop and mask to TEXAS 
sunshine_masked <- crop(sunshine, texas_counties_proj)
sunshine_masked <- terra::mask(sunshine_masked, texas_counties_proj)

# Check to see if it worked 
tm_shape(sunshine_masked) + 
  tm_raster() + 
  tm_shape(texas_counties) + 
  tm_borders()

```

Turn rasters into county polygons values 
```{r}
# Extract rasters into county polygons 
texas_counties$ghi <- exact_extract(sunshine_masked, texas_counties, 'mean')

# Make sunshine dataframe 
sunshine_df <- data.frame(countyfp =texas_counties$countyfp, 
           cnty_name = texas_counties$cnty_name,
           namelsad = texas_counties$namelsad,
           ghi = texas_counties$ghi 
           #geometry = texas_counties$geometry
           )

#sunshine_df <- st_as_sf(sunshine_df)

```

## Coal Deposites Data 

County names are all caps 
```{r}
# Make county names lowercase, expect the first one 
coal_dep$cnty_name <- str_to_title(coal_dep$county)
coal_dep <- coal_dep %>% select(-c(county))
```


## Wind Data 

Crop and mask to Texas 
```{r}
# Look at the extent of the wind data 
# Data encompasses the USA 
#tm_shape(wind_envir80) + 
  #tm_raster()


# Make CRS match 
texas_counties_proj <- st_transform(texas_counties, crs(wind_envir80))

# Crop and mask to TEXAS 
# 80m 
wind_envir80_masked <- crop(wind_envir80, texas_counties_proj)
wind_envir80_masked <- terra::mask(wind_envir80_masked, texas_counties_proj)
# 100m 
wind_envir100_masked <- crop(wind_envir100, texas_counties_proj)
wind_envir100_masked <- terra::mask(wind_envir100_masked, texas_counties_proj)
# 120m 
wind_envir120_masked <- crop(wind_envir120, texas_counties_proj)
wind_envir120_masked <- terra::mask(wind_envir120_masked, texas_counties_proj)
# 140m 
wind_envir140_masked <- crop(wind_envir140, texas_counties_proj)
wind_envir140_masked <- terra::mask(wind_envir140_masked, texas_counties_proj)

# Check to see if it worked 
# tm_shape(wind_envir80_masked) + 
#   tm_raster() + 
#   tm_shape(texas_counties) + 
#   tm_borders()
# 
# tm_shape(wind_envir100_masked) + 
#   tm_raster() + 
#   tm_shape(texas_counties) + 
#   tm_borders()
# 
# tm_shape(wind_envir120_masked) + 
#   tm_raster() + 
#   tm_shape(texas_counties) + 
#   tm_borders()
# 
# tm_shape(wind_envir140_masked) + 
#   tm_raster() + 
#   tm_shape(texas_counties) + 
#   tm_borders()

```

Turn rasters into county polygons values 
```{r}
texas_counties$wind_80m <- exact_extract(wind_envir80_masked, texas_counties, 'mean')
texas_counties$wind_100m <- exact_extract(wind_envir100_masked, texas_counties, 'mean')
texas_counties$wind_120m <- exact_extract(wind_envir120_masked, texas_counties, 'mean')
texas_counties$wind_140m <- exact_extract(wind_envir140_masked, texas_counties, 'mean')

# Make sunshine dataframe 
wind_texas <- data.frame(
  countyfp = texas_counties$countyfp,
  cnty_name = texas_counties$cnty_name,
  namelsad = texas_counties$namelsad,
  wind_80m = texas_counties$wind_80m, 
  wind_100m = texas_counties$wind_100m, 
  wind_120m = texas_counties$wind_120m, 
  wind_140m = texas_counties$wind_140m 
  #geometry = texas_counties$geometry
)

#wind_texas <- st_as_sf(wind_texas)

```

Remove environmental factors from `texas_counties`
```{r}
texas_counties <- texas_counties %>% 
  select(-c(ghi, wind_80m, wind_100m, wind_120m, wind_140m))

```

# Save data 

As CSVs 
```{r}
# Environmental Factors 
write.csv(coal_dep, file = here::here("data", "processed", "coal_deposites_texas.csv"))
write.csv(sunshine_df, file = here::here("data", "processed", "ghi_texas.csv"))
write.csv(wind_texas, file = here::here("data", "processed", "wind_texas.csv"))

# Socio demographics 
write.csv(socio_demo_texas, file = here::here("data", "processed", "socio_demo_texas.csv"))

# Counties 
write.csv(texas_counties, file = here::here("data", "processed", "texas_counties.csv"))

# Energy 
write.csv(generator_TX, file = here::here("data", "processed", "generator_TX.csv"))
write.csv(multifuel_TX, file = here::here("data", "processed", "multifuel_TX.csv"))
write.csv(plant_TX, file = here::here("data", "processed", "plant_TX.csv"))
write.csv(solar_energy_TX, file = here::here("data", "processed", "solar_energy_TX.csv"))
write.csv(storge_TX, file = here::here("data", "processed", "storge_TX.csv"))
write.csv(utility_TX, file = here::here("data", "processed", "utility_TX.csv"))
write.csv(wind_energy_TX, file = here::here("data", "processed", "wind_energy_TX.csv"))

```

As parquet file 
```{r}
# Environmental factors 
write_parquet(coal_dep, sink = here::here("data", "processed", "coal_deposites_texas.parquet"))
write_parquet(sunshine_df, sink = here::here("data", "processed", "ghi_texas.parquet"))
write_parquet(wind_texas, sink = here::here("data", "processed", "wind_texas.parquet"))

# Socio demographics 
st_write_parquet(socio_demo_texas, dsn = here::here("data", "processed", "socio_demo_texas_sf.parquet"))

# Counties 
st_write_parquet(texas_counties, dsn = here::here("data", "processed", "texas_counties_sf.parquet"))

# Energy 
write_parquet(generator_TX, sink = here::here("data", "processed", "generator_TX.parquet"))
write_parquet(multifuel_TX, sink = here::here("data", "processed", "multifuel_TX.parquet"))
write_parquet(plant_TX, sink = here::here("data", "processed", "plant_TX.parquet"))
write_parquet(solar_energy_TX, sink = here::here("data", "processed", "solar_energy_TX.parquet"))
write_parquet(storge_TX, sink = here::here("data", "processed", "storage_TX.parquet"))
write_parquet(utility_TX, sink = here::here("data", "processed", "utility_TX.parquet"))
write_parquet(wind_energy_TX, sink = here::here("data", "processed", "wind_energy_TX.parquet"))

```



