Load Libraries 
```{r}
library(tidyverse)
library(dplyr)
library(sf)
library(openxlsx)
library(janitor)
library(htmlwidgets)
library(plotly)
library(tmap)
library(terra)
library(exactextractr)
```

Load Energy Data & Filter for TEXAS 
```{r}
utility_TX <- read.xlsx(here::here("data", "raw", "eia8602024", "1___Utility_Y2024.xlsx"), startRow = 2) %>% 
  clean_names() %>% 
  filter(state == "TX")

plant_TX <- read.xlsx(here::here("data", "raw", "eia8602024", "2___Plant_Y2024.xlsx"), startRow = 2) %>% 
  clean_names() %>% 
  filter(state == "TX")

wind_energy_TX <- read.xlsx(here::here("data", "raw", "eia8602024", "3_2_Wind_Y2024.xlsx"), startRow = 2) %>% 
  clean_names() %>% 
  filter(state == "TX")

solar_energy_TX <- read.xlsx(here::here("data", "raw", "eia8602024", "3_3_Solar_Y2024.xlsx"), startRow = 2) %>% 
  clean_names() %>% 
  filter(state == "TX")

storge_TX <- read.xlsx(here::here("data", "raw", "eia8602024", "3_4_Energy_Storage_Y2024.xlsx"), startRow = 2) %>% 
  clean_names() %>% 
  filter(state == "TX")

generator_TX <- read.xlsx(here::here("data", "raw", "eia8602024", "3_1_Generator_Y2024.xlsx"), startRow = 2) %>% 
  clean_names() %>% 
  filter(state == "TX")

multifuel_TX <- read.xlsx(here::here("data", "raw", "eia8602024", "3_5_Multifuel_Y2024.xlsx"), startRow = 2) %>% 
  clean_names() %>% 
  filter(state == "TX")
```

Load Texas Socio-demographic 
```{r}
#st_layers(here::here("data", "raw", "ACS_2019_5YR_TRACT_48_TEXAS.gdb"))

# Geom 
geom_texas <- st_read(here::here("data", "raw", "ACS_2019_5YR_TRACT_48_TEXAS.gdb"), layer = "ACS_2019_5YR_TRACT_48_TEXAS")%>% clean_names()

# Socio-demo info 
age_sex_texas <- st_read(here::here("data", "raw", "ACS_2019_5YR_TRACT_48_TEXAS.gdb"), layer = "X01_AGE_AND_SEX")%>% clean_names()
income_texas <- st_read(here::here("data", "raw", "ACS_2019_5YR_TRACT_48_TEXAS.gdb"), layer = "X19_INCOME")%>% clean_names()
race_texas <- st_read(here::here("data", "raw", "ACS_2019_5YR_TRACT_48_TEXAS.gdb"), layer = "X02_RACE")%>% clean_names()
poverty_texas <- st_read(here::here("data", "raw", "ACS_2019_5YR_TRACT_48_TEXAS.gdb"), layer = "X17_POVERTY")%>% clean_names()
education_texas <- st_read(here::here("data", "raw", "ACS_2019_5YR_TRACT_48_TEXAS.gdb"), layer = "X15_EDUCATIONAL_ATTAINMENT")%>% clean_names()

# Meta data 
metadata_socio_texas <- st_read(here::here("data", "raw", "ACS_2019_5YR_TRACT_48_TEXAS.gdb"), layer = "TRACT_METADATA_2019")
```

Load Texas counties 
```{r}
#texas_counties <- st_read(here::here("data", "raw", "Texas_County_Boundaries"))%>% clean_names() # Not what I want 
texas_counties <- st_read(here::here("data", "raw", "tl_2024_us_county"))%>% clean_names() %>% 
  filter(statefp == '48')# THIS IS WHAT I WANT 

```

Load Texas environmental 
```{r}
sunshine <- terra::rast(here::here("data", "raw", "nsrdbv3_ghi", "Annual GHI", "nsrdb3_ghi.tif"))%>% clean_names()
```

# Wrangling 

Making Socio-demographic dataframe spatial 
Join Socio-demographic dataframe  with `geom_texas` to add the countyfp column and make spatial 
```{r}
income_texas_sf <- dplyr::left_join(geom_texas, income_texas, by = c("geoid_data"= "geoid")) 
race_texas_sf <- dplyr::left_join(geom_texas, race_texas, by = c("geoid_data"= "geoid")) 
poverty_texas_sf <- dplyr::left_join(geom_texas, poverty_texas, by = c("geoid_data"= "geoid"))
education_texas_sf <- dplyr::left_join(geom_texas, education_texas, by = c("geoid_data"= "geoid"))
age_sex_texas_sf <- dplyr::left_join(geom_texas, age_sex_texas, by = c("geoid_data"= "geoid"))

# Remove non spatial dfs 
#rm(list = c('income_texas', 'race_texas', 'poverty_texas', 'education_texas', 'age_sex_texas'))
```

Per county 
Pull out important columns 
```{r}
income_texas_county <- income_texas_sf %>% 
  group_by(countyfp) %>% 
  summarise(income_avg = mean(b19001e1, na.rm = TRUE))

race_texas_county <- race_texas_sf %>% 
  group_by(countyfp) %>% 
  summarise(pop_total = sum(b02001e1, na.rm = TRUE), 
            white_total = sum(b02001e2, na.rm = TRUE), 
            black_aa_total = sum(b02001e3, na.rm = TRUE), # Blacks / African American 
            ai_an_total = sum(b02001e4, na.rm = TRUE), # American India + Alaska Native
            asian_total = sum(b02001e5, na.rm = TRUE), 
            ah_pi_total = sum(b02001e6, na.rm = TRUE), # Native Hawian, other pacific islander
            other_total = sum(b02001e7, na.rm = TRUE),
            `2plus_total` = sum(b02001e8, na.rm = TRUE)) # two or more races 

poverty_texas_county <- poverty_texas_sf %>% 
  group_by(countyfp) %>% 
  summarise(hh_pov_status_total = sum(b17017e1, na.rm = TRUE))# households Counts in poverty 


education_texas_county <- education_texas_sf %>% 
  group_by(countyfp) %>% 
  summarise(edu_total = sum(b15001e1, na.rm = TRUE))#  EDUCATIONAL ATTAINMENT FOR THE POPULATION 18 YEARS AND OVER 


age_sex_texas_county <- age_sex_texas_sf %>% 
  group_by(countyfp) %>% 
  summarise(male_total = sum(b01001e2, na.rm = TRUE), 
            female_total = sum(b01001e26, na.rm = TRUE), 
            median_age_mean = mean(b01002e1, na.rm = TRUE), 
            median_male_age_mean = mean(b01002e2, na.rm = TRUE), 
            median_female_age_mean = mean(b01002e3, na.rm = TRUE))
  
```


Needs to be per-capita data 
Merge all datasets together 
```{r}
# Make the COUNTYFP a numeric and order the lists based on it 
race_texas_county <- race_texas_county %>% arrange(as.numeric(countyfp))
income_texas_county <- income_texas_county %>% arrange(as.numeric(countyfp))
poverty_texas_county <- poverty_texas_county %>% arrange(as.numeric(countyfp))
education_texas_county <- education_texas_county %>% arrange(as.numeric(countyfp))
age_sex_texas_county <- age_sex_texas_county %>% arrange(as.numeric(countyfp))


# Merge the sociodemographic data 
final_df_sf <- cbind(race_texas_county, income_texas_county, poverty_texas_county, education_texas_county, age_sex_texas_county) 

# Remove unnessary columns 
final_df_sf <- final_df_sf %>% 
  select(-c(`countyfp.1`, `countyfp.2`, `countyfp.3`, `countyfp.4`, `Shape.1`, `Shape.2`, `Shape.3`, `Shape.4`))

```

Pre capita 
```{r}
# Divide the columns by the population column 
final_df_sf[, 3:17] <- final_df_sf[, 3:17] / final_df$pop_total
```

Add the county names to the data 
```{r}
# Drop the geometry from the socio demographic data 
final_df <- final_df_sf %>% 
  st_drop_geometry()

# Join with 
socio_demo_texas <- left_join(final_df, texas_counties, by = "countyfp")

# Make spatail 
socio_demo_texas <- st_as_sf(socio_demo_texas)
```


Looking at outputs 
`socio_demo_texas` should be 254 rows (254 counties in TX)
```{r}
dim(socio_demo_texas)
dim(texas_counties)
dim(final_df)
n_distinct(socio_demo_texas$countyfp)
```
Create map to show results and dataframes 
```{r}
TX_counties_static <- ggplot(texas_counties) + 
  geom_sf(aes(text = paste0("County name:", {name}, "<br>", 
                            "countyfp:", {countyfp}, "<br>",
                            "geoid:", {geoid}, "<br>"
                            )))

final_df_static <- ggplot(final_df_sf) + 
  geom_sf(aes(text = paste0("countyfp:", {countyfp}, "<br>"
                            )))


socio_demo_static <- ggplot(socio_demo_texas) + 
  geom_sf(aes(text = paste0("County name:", {name}, "<br>", 
                            "countyfp:", {countyfp}, "<br>" )))

TX_counties_static
final_df_static
socio_demo_static

ggplotly(TX_counties_static, tooltip = "text")
ggplotly(final_df_static, tooltip = "text")
ggplotly(socio_demo_static, tooltip = "text")
```


```{r}
socio_demo <- ggplot(socio_demo_texas) + 
  geom_sf(aes(fill = ai_an_total,
              text = paste0("County name:", {name}, "<br>", 
                            "countyfp:", {countyfp}, "<br>",
                            "geoid:", {geoid}, "<br>"
                            )))

ggplotly(socio_demo, tooltip = "text")

```

Sunshine Wrangling 
```{r}
# Its South and North America
tm_shape(sunshine) + 
  tm_raster()

# Crop and mask to TEXAS 
texas_bbox <- ext(texas_counties) # Extracts the bounding box from the vector geometry
sunshine_masked <- crop(sunshine, texas_bbox)
sunshine_masked <- terra::mask(sunshine_masked, texas_counties)

# Check to see if it worked 
tm_shape(sunshine_masked) + 
  tm_raster() + 
  tm_shape(texas_counties) + 
  tm_borders()

sunshine_polygons <- as.polygons(sunshine_masked)

tm_shape(sunshine_polygons) + 
  tm_polygons() 

# Extract rasters into county polygons 
texas_counties$ghi <- exact_extract(sunshine_masked, texas_counties, 'mean')


```


