---
title: "data_viz_prep"
format: html
---
# OVERARCHING QUESTION:
Do Texas communities optimize their energy generation around local environmental resources, and do socioeconomic and demographic disparities influence how those resources are utilized?

```{r}
#install.packages('DBI')
# Libraries
library(tidyverse)
library(DBI)
library(dbplyr)
library(duckdb)

# Connection
conn <- DBI::dbConnect(duckdb::duckdb(),
                       dbdir = "database_finalProj.db")

# Tables 
DBI::dbListTables(conn)
```

## **QUESTION 1:** ENVIR ALIGNMENT - Do Texas communities leverage local environmental resources in their energy generation mix?--

### Q1.A: Is solar energy adoption associated with solar irradiance levels across Texas communities?

```{r}
solar_energy <- tbl(conn, "solar_energy")
solar_energy
```

**Sunshine**
Need to later - JOIN texas_counties polygons 
```{r}
# SQL Query 
solar_energy_texas <- DBI::dbGetQuery(
  conn,
  "SELECT cnty_name, COUNT(*) AS count ,AVG(nameplate_capacity_mw) AS capacity_avg, AVG(ghi) AS ghi_avg FROM solar_energy AS solar
JOIN ghi USING(cnty_name)
GROUP BY cnty_name
ORDER BY cnty_name"
)


# Scatter 
ggplot(solar_energy_texas) + 
  geom_point(aes(y = capacity_avg, 
                 x = ghi_avg)) + 
  geom_smooth(aes(y = capacity_avg, 
                 x = ghi_avg), 
              method = lm, se = FALSE) + 
  theme_classic()
  

```


###  Q1.B: Is wind energy capacity associated with wind resource availability across Texas communities?

```{r}
wind_energy <- tbl(conn, "wind_energy")
wind_energy
```

**Wind**
```{r}
# SQL Query 
wind_energy_texas <- DBI::dbGetQuery(
  conn,
  "SELECT cnty_name, AVG(wind_100m) AS windspeed_100m_avg, AVG(nameplate_capacity_mw) AS capacity_avg FROM wind_energy
    JOIN wind USING(cnty_name)
    GROUP BY cnty_name
    ORDER BY cnty_name"
)

wind_energy_texas

# Scatter 
ggplot(wind_energy_texas) + 
  geom_point(aes(y = capacity_avg, 
                 x = windspeed_100m_avg)) + 
  geom_smooth(aes(y = capacity_avg, 
                 x = windspeed_100m_avg), 
              method = lm, se = FALSE) + 
  theme_classic()
  
```


## QUESTION 2: How is energy generation distributed across socioeconomically disadvantaged communities in Texas?

### Q2.A: How does residential energy source distribution vary across income brackets? 

```{r}
socio_demo <- tbl(conn, "socio_demo")
socio_demo
```

```{r}
income_energy <- dbGetQuery(
  conn,
  "SELECT cnty_name,
        AVG(income_avg) AS avg_income,
        SUM(total_voltage) AS total_voltage,
        COUNT(*) AS plant_count,
        AVG(solar.nameplate_capacity_mw) AS solar_avg_capacity_mw,
        AVG(wind.nameplate_capacity_mw) AS wind_avg_capacity_mw
    FROM plant_energy
    -- Join the other datasets
    JOIN socio_demo USING (cnty_name)
    JOIN solar_energy AS solar USING (cnty_name)
    JOIN wind_energy AS wind USING (cnty_name)
    GROUP BY cnty_name
    ORDER BY cnty_name"
) 

income_energy

```

```{r}

income_energy <- income_energy %>% 
  mutate(avg_income_bin = case_when(
    income_energy$avg_income < 25000 ~ "Under 25K", 
    income_energy$avg_income < 50000 ~ "25K -50K", 
    income_energy$avg_income < 100000 ~ "50K - 100K", 
    income_energy$avg_income < 150000 ~ "100K - 150K", 
    income_energy$avg_income < 150000 ~ "150K+" 
    ))

income_energy %>% 
  group_by(avg_income_bin) %>% 
  summarise(across(everything(), mean, na.rm = TRUE))

income_energy


```



# Close databse 
```{r}
# CLOSE DATABASE CONNECTION  - MUST DO 
DBI::dbDisconnect(conn, shutdown = TRUE)

```

